A Quantitative Genetic Linked Selection Model of the Negative Selection Process in Humans

### 2023-07-18 manuscript/manuscript.tex
ms updates

    Across the human genome, there are large-scale fluctuations in genetic
    diversity caused by the indirect effects of selection. This can be thought
    of as a ``linked selection signal" that reflects selection's effects
    varying according to the placement of functional regions and recombination
    rates along the genome. Previous work has shown that negative selection
    against the steady influx of new deleterious mutations into conserved
    regions is the predominant mode of selection in humans. However, the
    theoretic model that underpins these results, classic Background Selection
    theory, is only applicable when new mutations are so deleterious that they
    cannot drift up in frequency and ultimately fix in the population. Here, we
    develop an alternate theoretic basis for genome-wide linked selection based
    on a quantitative genetics view, which considers how diversity is reduced
    according to the arrangement of additive fitness variance along the genome.
    Using a recent model that jointly predicts the equilibrium fitness variance
    and substitution rates due to deleterious mutations, we estimate the
    distribution of selection coefficients and mutation rate across three human
    populations. While our model can accommodate weaker selection, we find
    consistent evidence across all populations of very strong selection against
    deleterious mutations. We also find that our model's predicted substitution
    rates are qualitatively consistent with observed divergences in features
    under weak selective constraint like introns and UTRs, but less so in
    regions under strong selective constraint. While our model explains up to
    67\% of the out-sample variance in diversity at the megabase scale, we find
    evidence of residual strong selection against loss-of-function variants
    unfit by our model, indicating the dominant role negative selection plays
    in the human genome. Altogether, our genome-wide model of the linked
    selection is a step towards uniting evolutionary quantitative genetics with
    substitution processes.

@@

The continual influx of new mutations into populations is the ultimate source
of all adaptations, but the vast majority of mutations either do not affect
fitness or are deleterious. Negative selection (i.e. purifying selection) works
to eliminate these deleterious mutations from the population, thus we expect
them to appear at low frequencies within populations
\parencite{Haldane1927-ga}, and less likely to fix between lineages. The best
predictor of whether a new mutation will reduce fitness is if it occurs in a
functional region of the genome, which are often conserved over deep
phylogenetic timescales \parencite{Siepel2005-wh,Margulies2003-sn}. Since these
conserved regions reflect the product of millions of years of evolutionary
optimization, it follows that deleterious variants should be the overwhelming
majority of segregating variation in these regions. Furthermore, rare variation
in these regions is responsible for a significant portion of phenotypic
variation and disease in humans
\parencite{Zeng2018-ci,Lek2016-cb,Karczewski2020-ky,Tennessen2012-ge}.

Selection on both beneficial and deleterious variants perturbs the allele
frequencies of neighboring linked sites, a phenomenon known as linked selection
\parencite{Nordborg1996-nq,Maynard_Smith1974-zr,Barton1998-aj,Charlesworth1993-gb}.
Since deleterious variation is clustered in functional portions of the genome,
we expect linked selection to reduce levels of diversity around conserved
segments (e.g. genes, regulatory elements, etc.). The genomic arrangement of
these evolutionarily conserved regions and recombination rates create a
large-scale spatial signal of the linked selection process in genetic diversity
along the genome. Since genome-wide recombination maps and putatively conserved
segments are available for many species, there has been consistent effort to
fit models to this linked selection signal to estimate population genetic
parameters such as the strength of selection and the deleterious mutation rate
\parencite{Hudson1995-pt,McVicker2009-ax}, as well as distinguish and quantify
the roles of positive and negative selection and estimate the rate of
beneficial mutations \parencite{Elyashiv2016-vt,Murphy2022-sj}. In humans,
previous work has shown that negative selection plays the dominant role in
shaping megabase-scale patterns of diversity, with positive selection playing a
nearly negligible role \parencite{Murphy2022-sj}.

Prior work to model the reduction in linked diversity due to deleterious
mutations has largely relied on the classic Background Selection (BGS) model
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1995-xc}.
While this model has been successful in explaining many patterns of diversity,
some its simplifying assumptions may distort inferences about the negative
selection process. First, since fixation probabilities ultimately depend on the
product of the deleterious selection coefficient (s) and population size (N),
the efficacy of negative selection should depend on demography. Unfortunately,
accommodating demography into theoretic negative selection models remains an
open, difficult problem \parencite{Zeng2013-ep,Johri2020-oj}. Second, building
off classic models of mutation-selection balance
\parencite{Crow1970-wm,Kimura1966-bk}, the BGS model assumes that new mutations
are sufficiently deleterious that they are invariably driven to loss. Under
this assumption, the effect of selection is well-approximated by simply
rescaling the neutral coalescent by a reduction factor known as $B =
\nicefrac{N}{N_e}$ \parencite{Charlesworth2013-kl}. However, this simple
rescaling approach is not appropriate across parts of parameter space that are
relevant to natural populations \parencite{McVean2000-bt,Good2014-yz}. In
particular, the BGS model cannot accommodate the possibility of fixation for
mutations with $Ns \ll 1$, diversity is incorrectly predicted to fall to zero
as the strength of selection diminishes. This weak selection problem is related
to the wider problem of modeling the substitution rate of deleterious mutation
(i.e. the “ratchet” rate) and incorporating the impact of perturbations from
selective interference.

In this work, we extend another class of linked selection models that derive
from quantitative genetics to address limitations of the classic BGS model.
These models express linked selection in terms of how polygenic fitness
variance creates weak genetic draft, and which increases pairwise coalescence
rates. While these models can theoretically accommodate additive fitness
variance from any source as long as its rate of change is not too rapid, we
focus specifically on a deleterious-mutations-only model of fitness variance
from \textcite{Santiago2016-mu}. This model is identical to BGS when selection
against deleterious mutations is strong, but it also correctly predicts the
reduction in diversity when selection is weak by jointly predicting the
deleterious substitution rate. We extend the Santiago and Caballero (hereafter
the SC16) model of the negative selection process so that it can be fit to
patterns of genome-wide diversity, according to the spatial distribution of
genomic features that could harbor deleterious fitness variation. Using forward
simulations, we show this model leads to more accurate estimates of the
distribution of fitness effects under weak selection. Finally, we apply our
composite-likelihood method to human population genomic data and provide new
estimates of the negative selection process in humans.

@@

Similar to previous work, we aim to characterize the negative selection process
through the signal left on nucleotide diversity at linked sites. Although
nucleotide diversity is a less-informative statistic than the full site
frequency spectrum, it is much easier to derive theoretic models of how linked
selection alters expected pairwise coalescence times than to describe the
entire genealogy. Still, inferences drawn from the linked selection signal
about selective processes are only as accurate as the underlying theoretic
model. Our work is motivated by a well-known limitation of classic background
selection theory: it only accurately predicts levels of linked diversity when
new mutations are strongly deleterious
\parencite{Charlesworth1993-gb,McVean2000-bt,Good2013-lp,Gordo2002-dr}.
Unfortunately, modifying background selection theory so that it accommodates
weak selection has been challenging problem
\parencite{Good2014-yz,Haigh1978-gt,Higgs1995-xc}. Here, we develop intuition
for why the classic BGS model works well for strong selection but not weak, and
how formulating selection in terms of the additive fitness variance it creates
can alleviate these problems.

Throughout, we imagine a negative selection process where a continual influx of
deleterious mutations occur at a rate of $\mu$ per basepair per generation in a
conserved region of $L$ basepairs, such that the region-wide per generation
mutation rate is $U = \mu L$. Each mutation imposes a selective cost of $s$ in
heterozygotes and $2s$ in homozygotes, and fitness effects are multiplicative
across sites. When selection against a new mutation is strong, the stochastic
perturbations it experiences due to random genetic drift are weak relative to
the frequency changes caused by selection. In this case, every deleterious
mutation is destined to eventual extinction, and only the fraction of lineages
without any mutations contribute to the long-run ancestry of future
populations. The classic background selection approximates this process by
rescaling the neutral coalescent with an effective population size of $N_0 = N
\exp(-\nicefrac{U}{s})$
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}.
This approximation works well for two reasons (though see
\cite{Cvijovic2018-vd,Walczak2012-fi,Nicolaisen2012-vs}). First, the number of
mutations per haplotype has has a stationary Poisson distribution with an
average equal to the deterministic mutation-selection equilibrium
\parencite{Haldane1927-ga}. Second, the time it takes for a present-day lineage
to trace their ancestry back to a mutation-free ancestor (the ``delay phase")
is negligible compared to the coalescence timescale of $N_0$ among
mutation-free ancestors (\cite{Durrett2008-ql}, p. 213; \cite{Good2014-yz}).

By contrast, when selection is weak, the frequency trajectories of both neutral
and selected sites are strongly impacted by stochastic perturbations from two
sources. First, genetic drift is due to Mendelian segregation and non-heritable
fitness differences; the magnitude of its perturbations are set by the
\emph{drift-effective} population size, $N_e$. Second, genetic draft occurs
when an allele becomes randomly associated with a fitness background that
perturbs its trajectory across generations \parencite{Neher2013-dz}. Like
genetic drift, the perturbations due to draft are directionless but overall
increase the variance in allele frequency change. Unlike drift, the impact of
the stochastic perturbations due to draft are autocorrelated and build up over
generations until the fitness background recombines off
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. The stochastic
perturbations due to selection also impact selected sites, and alter their
dynamics, creating Hill--Robertson interference \parencite{Hill1966-kd}. Strong
draft can generate multiple-merger coalescences that create discontinuities in
allele frequency trajectories, making diffusion approximations ill-suited
\parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}. However, when selection
is weak, draft is well-approximated as an increase in the rate of drift (or
equivalently, a reduction in effective population size).

@@

We can define the draft-effective population size $N_d$ by including the total
additional variance created by heritable fitness in Wright's
(\citeyear{Wright1938-tv}) equation for effective population size,

@@

We next validated our predicted B' reduction maps with realistic
chromosome-scale forward simulations of negative selection in conserved
regions. Averaging over $100$ replicates, we estimated the simulation empirical
reduction map and compared this to the predicted B and B' maps for the same
parameters. We find that our B' maps and the classic BGS theory B maps closely
match simulations when selection is strong (top row of Figure
\ref{fig:figure-1}E). This provides the first realistic forward-time
chromosome-scale simulation confirmation of classic BGS theory. However, we
find slight discrepancies in regions with low recombination (Figure
\ref{fig:figure-1}E). Second, we find our theory is vastly more accurate than
the classic B maps when selection is very weak ($2N_e s \ll 1$; bottom row of
Figure \ref{fig:figure-1}E), which is expected since BGS theory does not work
in this domain. Across all mutation and selection parameters simulated, the
relative error of the classic B maps is 14.6\% whereas the relative error in
the new B' maps is 5\%. Additionally, we find that the mean squared error
between simulations and B' maps is close to the theoretic lower bound set by
the coalescence variance \parencite[see Methods XXX]{Tajima1983-gu}.

@@

Our predictions are qualitatively consistent with the observed divergence along
the human lineage (Figure \ref{fig:figure-7}). As expected, the predicted
substitution rate in features unlikely to be under selective constraint is very
close to the mutation rate. Throughout, we report our substitution rates as a
percent relative to the total mutation rate, $\mu$ (here fixed to $1.5 \times
10^{-8}$). In our PhastCons Priority model, introns and UTRs are estimated to
have have substitution rates that are $93.6\%$ of the mutation rate, and the
``other" feature is estimated at $97.8\%$.  When we use the constrained grid,
these change to $97.85\%$ and $99.98\%$ respectively. In our PhyloFit
estimates, the divergence rate in introns is $92.44\%$, 5' UTRs $75.5\%$, and
3' UTRs $72.65\%$. By contrast, our predicted substitution rates for PhastCons
regions are much lower: $0.48\%$ for the default grid, and $1.65\%$ for the
constrained grid. For comparison, the PhyloFit estimates for PhastCons regions
is much higher, at $51.23\%$. Since this would include beneficial substitutions
that would not be predicted by our model, this suggests a very high proportion
of beneficial substitutions XXX check

>>>>

    Across the human genome, there are large-scale fluctuations in genetic
    diversity caused by the indirect effects of selection. This can be thought
    of as a ``linked selection signal" that reflects selection's effects
    varying according to the placement of functional regions and recombination
    rates along the genome. Previous work has shown that negative selection
    against the steady influx of new deleterious mutations into conserved
    regions is the predominant mode of selection in humans. However, the
    theoretic model that underpins these results, classic Background Selection
    theory, is only applicable when new mutations are so deleterious that they
    cannot drift up in frequency and ultimately fix in the population. Here, we
    develop an alternate theoretic basis for genome-wide linked selection based
    on a quantitative genetics view, which considers how diversity is reduced
    according to the arrangement of additive fitness variance along the genome.
    Using a recent model that jointly predicts the equilibrium fitness variance
    and substitution rates due to deleterious mutations, we estimate the
    distribution of selection coefficients and mutation rate across three human
    populations. While our model can accommodate weaker selection, we find
    consistent evidence across all populations of very strong selection against
    deleterious mutations. We also find that our model's predicted substitution
    rates are qualitatively consistent with observed divergences in features
    under weak selective constraint like introns and UTRs, but less so in
    regions under strong selective constraint. While our model explains up to
    67\% of the out-sample variance in diversity at the megabase scale, we find
    evidence of residual strong selection against loss-of-function variants
    unfit by our model, indicating the dominant role negative selection plays
    in the human genome. Altogether, our genome-wide model of the linked
    selection is a step towards uniting population and quantitative genetic
    models of selection with the substitution processes.

@@

The continual influx of new mutations into populations is the ultimate source
of all adaptations, but the vast majority of mutations either do not affect
fitness or are deleterious. Negative selection (i.e. purifying selection) works
to eliminate these deleterious mutations from the population; thus we expect
them to appear at low frequencies within populations
\parencite{Haldane1927-ga}, and be less likely to fix between lineages. The
best predictor of whether a new mutation will reduce fitness is if it occurs in
a functional region of the genome, which are often conserved over deep
phylogenetic timescales \parencite{Siepel2005-wh,Margulies2003-sn}. Since these
conserved regions reflect the product of millions of years of evolutionary
optimization, it follows that deleterious variants should be the overwhelming
majority of segregating variation in these regions. Furthermore, rare variation
in these regions is responsible for a significant portion of phenotypic
variation and disease in humans
\parencite{Zeng2018-ci,Lek2016-cb,Karczewski2020-ky,Tennessen2012-ge}.

Selection on both beneficial and deleterious variants perturbs the allele
frequencies of neighboring linked sites, a phenomenon known as linked selection
\parencite{Nordborg1996-nq,Maynard_Smith1974-zr,Barton1998-aj,Charlesworth1993-gb}.
Since deleterious variation is clustered in functional portions of the genome,
we expect linked selection to reduce levels of diversity around conserved
segments (e.g. genes, regulatory elements, etc.). The genomic arrangement of
these evolutionarily conserved regions and recombination rates create a
large-scale spatial signal of the linked selection process in genetic diversity
along the genome. Since genome-wide recombination maps and putatively conserved
segments are available for many species, there has been consistent effort to
fit models to this linked selection signal to estimate population genetic
parameters such as the strength of selection and the deleterious mutation rate
\parencite{Hudson1995-pt,McVicker2009-ax}, as well as distinguish and quantify
the roles of positive and negative selection and estimate the rate of
beneficial mutations \parencite{Elyashiv2016-vt,Murphy2022-sj}. In humans,
previous work has shown that negative selection plays the dominant role in
shaping megabase-scale patterns of diversity, with positive selection having a
nearly negligible impact \parencite{Murphy2022-sj}.

Prior work to model the reduction in linked diversity due to deleterious
mutations has largely relied on the classic Background Selection (BGS) model
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1995-xc}.
While this model has been successful in explaining many patterns of diversity,
some its simplifying assumptions may distort inferences about the negative
selection process. First, since fixation probabilities ultimately depend on the
product of the deleterious selection coefficient ($s$) and population size
($N$), the efficacy of negative selection should depend on demography.
Unfortunately, accommodating demography into theoretic negative selection
models remains an open, difficult problem \parencite{Zeng2013-ep,Johri2020-oj}.
Second, building off classic models of mutation-selection balance
\parencite{Crow1970-wm,Kimura1966-bk}, the BGS model assumes that new mutations
are sufficiently deleterious that they are invariably driven to loss. Under
this assumption, the effect of selection is well-approximated by simply
rescaling the neutral coalescent by a reduction factor known as $B =
\nicefrac{N}{N_e}$ \parencite{Charlesworth2013-kl}. However, this simple
rescaling approach is not appropriate across parts of parameter space that are
relevant to natural populations \parencite{McVean2000-bt,Good2014-yz}. In
particular, the BGS model cannot accommodate the possibility of fixation for
mutations with $2Ns \ll 1$, leading diversity to be incorrectly predicted to
fall to zero as the strength of selection diminishes. This weak selection
problem is related to the wider problem of modeling the substitution rate of
deleterious mutation (i.e. the ``ratchet" rate) and incorporating the impact of
interference from other neighboring selected sites.

In this work, we use another class of linked selection models that derive from
quantitative genetics to address limitations of the classic BGS model. These
models express linked selection in terms of how polygenic fitness variance
creates weak genetic draft, and which increases pairwise coalescence rates.
While these models can theoretically accommodate additive fitness variance from
any source as long as its rate of change is not too rapid, we focus
specifically on a deleterious-mutations-only model of fitness variance from
\textcite{Santiago2016-mu}. This model is identical to BGS when selection
against deleterious mutations is strong, but it also correctly predicts the
reduction in diversity when selection is weak by jointly predicting the
deleterious substitution rate. We extend the Santiago and Caballero (hereafter
the SC16) model of the negative selection process so that it can be fit to
patterns of genome-wide diversity, according to the spatial distribution of
genomic features that could harbor deleterious fitness variation. Using forward
simulations, we show this model leads to more accurate estimates of the
distribution of fitness effects under weak selection. Finally, we apply our
composite-likelihood method to human population genomic data and provide new
estimates of the negative selection process in humans.

@@

Similar to previous work, we aim to characterize the negative selection process
through the signal left on nucleotide diversity at linked sites. Although
nucleotide diversity is a less-informative statistic than the full site
frequency spectrum, it is much easier to derive theoretic models of how linked
selection alters expected pairwise coalescence times than to describe the
entire genealogy. Still, inferences drawn from the linked selection signal
about selective processes are only as accurate as the underlying theoretic
model. Our work is motivated by a well-known limitation of classic background
selection theory: it only accurately predicts levels of linked diversity when
new mutations are strongly deleterious
\parencite{Charlesworth1993-gb,McVean2000-bt,Good2013-lp,Gordo2002-dr}.
Unfortunately, modifying background selection theory so that it accommodates
weak selection has been challenging problem
\parencite{Good2014-yz,Haigh1978-gt,Higgs1995-xc}. Here, we develop intuition
for why the classic BGS model works well for strong selection but not weak, and
how formulating the equilibrium between new deleterious mutations and negative
selection in terms of the additive fitness variance can alleviate these
problems.

Throughout, we imagine a negative selection process where a continual influx of
deleterious mutations occur at a rate of $\mu$ per basepair per generation in a
conserved region of $L$ basepairs, such that the region-wide per generation
diploid mutation rate is $U = 2 \mu L$. Each mutation imposes a selective cost
of $s$ in heterozygotes and $2s$ in homozygotes, and fitness effects are
multiplicative across sites. When selection against a new mutation is strong,
the stochastic perturbations it experiences due to random genetic drift are
weak relative to the frequency changes caused by selection. In this case, every
deleterious mutation is destined to eventual extinction, and only the fraction
of lineages without any mutations contribute to the long-run ancestry of future
populations. The classic background selection approximates this process by
rescaling the neutral coalescent with an effective population size of $N_0 = N
\exp(-\nicefrac{U}{s})$
\parencite{Charlesworth1993-gb,Nordborg1996-nq,Hudson1995-pt,Hudson1994-oh}.
This approximation works well for two reasons (though see
\cite{Cvijovic2018-vd,Walczak2012-fi,Nicolaisen2012-vs}). First, the number of
mutations per haplotype has has a stationary Poisson distribution with an
average equal to the deterministic mutation-selection equilibrium
\parencite{Haldane1927-ga}. Second, the time it takes for a present-day lineage
to trace their ancestry back to a mutation-free ancestor (the ``delay phase")
is negligible compared to the coalescence timescale of $N_0$ among
mutation-free ancestors (\cite{Durrett2008-ql}, p. 213; \cite{Good2014-yz}).

By contrast, when selection is weak, the frequency trajectories of both neutral
and selected sites are strongly impacted by stochastic perturbations from two
sources. First, genetic drift, which is due to Mendelian segregation and
non-heritable fitness differences; the magnitude of its perturbations are set
by the \emph{drift-effective} population size, $N_e$. Second, genetic draft
occurs when an allele becomes randomly associated with a fitness background
that perturbs its trajectory across generations
\parencite{Neher2013-dz,Gillespie2000-mh,Gillespie2001-mv}. Like genetic drift,
the perturbations due to draft are directionless but overall increase the
variance in allele frequency change. Unlike drift, the impact of the stochastic
perturbations due to draft are autocorrelated and build up over generations
until the fitness background recombines off
\parencite{Robertson1961-ho,Santiago1995-hx,Buffalo2019-qs}. The stochastic
perturbations due to selection also impact selected sites, and alter their
dynamics, creating Hill--Robertson interference \parencite{Hill1966-kd}. Strong
draft can generate multiple-merger coalescences that create discontinuities in
allele frequency trajectories, making diffusion approximations ill-suited
\parencite{Gillespie2000-mh,Der2011-it,Neher2013-dz}. However, when selection
is weak, draft is well-approximated as an increase in the rate of drift (or
equivalently, a reduction in effective population size).

@@

We define the draft-effective population size $N_d$ by including the total
additional variance created by heritable fitness in Wright's
(\citeyear{Wright1938-tv}) equation for effective population size,

@@

We validated our predicted B' reduction maps with realistic chromosome-scale
forward simulations of negative selection in conserved regions. Averaging over
$100$ replicates, we estimated the simulation empirical reduction map and
compared this to the predicted B and B' maps for the same parameters. We find
that our B' maps and the classic BGS theory B maps closely match simulations
when selection is strong (top row of Figure \ref{fig:figure-1}E). This provides
the first realistic forward-time chromosome-scale simulation confirmation of
classic BGS theory. However, we find slight discrepancies in regions with low
recombination (Figure \ref{fig:figure-1}E). Second, we find our theory is
vastly more accurate than the classic B maps when selection is very weak ($2N_e
s \ll 1$; bottom row of Figure \ref{fig:figure-1}E), which is expected since
BGS theory does not work in this domain. Across all mutation and selection
parameters simulated, the relative error of the classic B maps is 14.6\%
whereas the relative error in the new B' maps is 5\%. Additionally, we find
that the mean squared error between simulations and B' maps is close to the
theoretic lower bound set by the coalescence variance
\parencite{Tajima1983-gu}.

@@

We find that our predicted substitution rates are qualitatively consistent with
the observed divergence along the human lineage for all features except
PhastCons (Figure \ref{fig:figure-7}).

As expected, the predicted substitution
rate in features unlikely to be under selective constraint is very close to the
mutation rate. Throughout, we report our substitution rates as a percent
relative to the total mutation rate, $\mu$ (here fixed to $1.5 \times
10^{-8}$). In our PhastCons Priority model, introns and UTRs are estimated to
have have substitution rates that are $93.6\%$ of the mutation rate, and the
``other" feature is estimated at $97.8\%$.  When we use the constrained grid,
these change to $97.85\%$ and $99.98\%$ respectively. In our PhyloFit
estimates, the divergence rate in introns is $92.44\%$, 5' UTRs $75.5\%$, and
3' UTRs $72.65\%$. By contrast, our predicted substitution rates for PhastCons
regions are much lower: $0.48\%$ for the default grid, and $1.65\%$ for the
constrained grid. For comparison, the PhyloFit estimates for PhastCons regions
is much higher, at $51.23\%$. Since this would include beneficial substitutions
that would not be predicted by our model, this suggests a very high proportion
of beneficial substitutions XXX check
